---
title: "AD_MR"
author: "Alish Palmos"
date: "09/02/2021"
output: html_document
editor_options:
  chunk_output_type: console
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

#Set-up
```{r echo=FALSE}
rm(list=ls())
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

setwd("/scratch/groups/ukbiobank/usr/alish/AD_MR/")
```

```{r include=FALSE}
library(data.table)
library(jtools)
library(knitr)
library(broom)
library(sandwich)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(sgof)
library(TwoSampleMR)
library(qvalue)
library(kableExtra)
library(vroom)

```


```{r}
#get MR function
get_mr <- function(study, dir, extra = NULL){

  if(is.null(extra)){

  current_dir <- dir
  setwd(paste0(current_dir,"/",study,"_mr/output/"))

  list <- list.files(pattern = "*.gsmr")

  mr <- lapply(list, read.delim)

  bind <- do.call(rbind.data.frame, mr)

  title <- str_to_title(study)

  bind$study <- paste0(title)

  bind$file_name <- rep(list, each=2)

  bind <- bind %>%
    filter(nsnp >= 1)

  bind_n <- count(bind)

  assign(paste0(study,"_mr"),value = bind)

  }

  else

    current_dir <- dir

  setwd(paste0(current_dir,"/",study,"_mr/",extra,"output/"))

  list <- list.files(pattern = "*.gsmr")

  mr <- lapply(list, read.delim)

  bind <- do.call(rbind.data.frame, mr)

  title <- str_to_title(study)

  bind$study <- paste0(title)

  bind$file_name <- rep(list, each=2)

  bind <- bind %>%
    filter(nsnp >= 1)

  bind_n <- count(bind)

  assign(paste0(study,"_mr"),value = bind)

}

```


This RMarkdown shows the results from a MRWAS (Mendelian Randomization Wide Association Study).

All currently available GWAS summary statistics of inflammatory markers were amassed over the course a few months.


Covid GWAS used in this study:

Very severe respiratory confirmed covid vs. not hospitalized covid - October data release (https://www.covid19hg.org/results/)



Sun et al: https://pubmed.ncbi.nlm.nih.gov/29875488

Suhre et al: https://pubmed.ncbi.nlm.nih.gov/28240269

Wood et al: https://pubmed.ncbi.nlm.nih.gov/23696881

Sliz et al: https://www.ncbi.nlm.nih.gov/pubmed/?term=31217265

Folkerson et al: http://dx.plos.org/10.1371/journal.pgen.1006706

Ahola et al: https://pubmed.ncbi.nlm.nih.gov/27989323

Scallop et al: https://www.nature.com/articles/s42255-020-00287-2#Abs1

Bretherik et al: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7337286


Note that more studies are available for analyses. They're currently being cleaned and added as the data generation protocol progresses.


GSMR was used as the base MR method: https://www.nature.com/articles/s41467-017-02317-2


P-value threshold was set to 5E-6, due to lack of power in some GWAS. Thus, this study is exploratory.


Only comparisons where 10+ SNPs were in common between the exposure and the outcome were included. This is routine MR practice.


Bonferroni p-value threshold is stated below.

#PROTEIN -> AD
#Get all GSMR data

```{r echo=FALSE}

sun_mr <- get_mr(study = "sun", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
suhre_mr <- get_mr(study = "suhre", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
folk_mr <- get_mr(study = "folk", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
sliz_mr <- get_mr(study = "sliz", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
wood_mr <- get_mr(study = "wood", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
ahol_mr <- get_mr(study = "ahol", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
scal_mr <- get_mr(study = "scal", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
hill_mr <- get_mr(study = "hill", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
hogl_mr <- get_mr(study = "hogl", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
enroth_mr <- get_mr(study = "enroth", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")
deco_mr <- get_mr(study = "deco", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/")

```

# Combine all to count how many exposures we had as proteins
```{r echo=FALSE}
# combine all to count how many exposures and outcomes we had
mr_df_all <- rbind(folk_mr, sliz_mr, suhre_mr, sun_mr, wood_mr, ahol_mr, scal_mr, hill_mr, hogl_mr, enroth_mr, deco_mr)
als <- str_count(mr_df_all$Exposure, "AD_Schwartzentruber")
als[is.na(als)] <- 0
sum(als)
```

# Combine all to count how many exposures we had as AD
```{r}

markers <- mr_df_all[mr_df_all$Outcome %like% "AD_Schwartzentruber", ]
nrow(markers)

```

# Combine full data, work out odds ratios and 95% CI
```{r}

mr_df_all$Beta_exponent <- exp(mr_df_all$bxy)
mr_df_all$LCI <- mr_df_all$Beta_exponent - (mr_df_all$se * 1.96)
mr_df_all$UCI <- mr_df_all$Beta_exponent + (mr_df_all$se * 1.96)

mr_df_all$Exposure <- gsub(".txt", "",mr_df_all$Exposure)
mr_df_all$Outcome <- gsub(".txt", "",mr_df_all$Outcome)
```

# Compute q-values and add them to the large combined data frame
```{r}
pvalues <- mr_df_all$p
qobj <- qvalue(p = pvalues)
lfdr <- qobj$lfdr
summary(qobj)
hist(qobj)
plot(qobj)

qvalues <- as.matrix(qobj$qvalues)
mr_df_all$qvalue <- cbind(qvalues)
```

# Adjusted p-values using SGoF method
```{r}
p <- SGoF(mr_df_all$p)
summary(p)

plot(p)

p_list <- p$Adjusted.pvalues

mr_df_all <- mr_df_all[order(mr_df_all$p),]
mr_df_all$sgof <- p_list
```

# All effects AD outcome

```{r}

AD_outcome <- mr_df_all[mr_df_all$Outcome %like% "AD_Schwartzentruber", ]

  AD_outcome %>%
  kbl() %>%
  kable_classic_2(full_width = F)

```

# Sort by significant p-values and filter
```{r}
mr_df_filtered <- AD_outcome %>%
  filter(qvalue < 0.05)

mr_df_filtered <- mr_df_filtered %>%
  rename(Beta = bxy, StandardError = se, Pvalue = p, SNPs = nsnp, LowerCI = LCI, UpperCI =  UCI)

mr_df_filtered <- mr_df_filtered %>% arrange(desc(Beta_exponent))

mr_df_filtered$Log_Beta <- log(mr_df_filtered$Beta_exponent)
mr_df_filtered$Log_LowerCI <- log(mr_df_filtered$LowerCI)
mr_df_filtered$Log_UpperCI <- log(mr_df_filtered$UpperCI)

  mr_df_filtered %>%
  kbl() %>%
  kable_classic_2(full_width = F)
```

# Create a figure with AD as the outcome
```{r fig.width=20, fig.height=40, echo=FALSE}

mr_df_chart <- mr_df_filtered %>%
  filter(qvalue < 0.0001)

AD_outcome_stringent <- mr_df_chart
vroom_write(x = AD_outcome_stringent, file = "/scratch/groups/ukbiobank/usr/alish/AD_MR/AD_outcome_stringent")

mr_df_chart <- mr_df_chart[complete.cases(mr_df_chart), ]

#Tidy up the names of some proteins
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)

mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)
mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)
mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)
mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)

mr_df_chart <- mr_df_chart[!duplicated(mr_df_chart$Exposure), ]

library(scales)

ggplot(mr_df_chart, aes(x=reorder(as.factor(Exposure), Beta_exponent) , y=Beta_exponent)) +
    geom_boxplot(fill="slateblue") + coord_flip() + geom_errorbar(aes(ymax = UpperCI, ymin = LowerCI))+ ggtitle("Outcome - Alzheimer's") +
  xlab("Exposure Variable") + ylab("Odds Ratio (95% Confidence Interval)")+ geom_hline(yintercept=1, linetype="dashed", color = "red") +
  theme_bw() + theme(text=element_text(size=25, family="Arial")) +
  geom_hline(linetype = "dashed", yintercept = 1) +
    theme(panel.grid.major.x = element_line(size = 0.5,
                                        linetype = 'dashed',
                                        colour = "gray41"),
        panel.grid.major.y = element_line(size = 0.1,
                                        linetype = 'solid',
                                        colour = "gray62")) +
  scale_y_log10(breaks = c(-1, 0.5, 1, 1.5, 2, 2.5)) +
  guides(color = guide_legend(reverse = TRUE)) +
  coord_flip(clip = "off", ylim = c(-1,3))

```

# AD -> PROTEIN
#Get all GSMR data

```{r echo=FALSE}

sun_mr <- get_mr(study = "sun", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
suhre_mr <- get_mr(study = "suhre", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
folk_mr <- get_mr(study = "folk", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
sliz_mr <- get_mr(study = "sliz", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
wood_mr <- get_mr(study = "wood", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
ahol_mr <- get_mr(study = "ahol", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
scal_mr <- get_mr(study = "scal", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
hill_mr <- get_mr(study = "hill", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
hogl_mr <- get_mr(study = "hogl", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
enroth_mr <- get_mr(study = "enroth", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")
deco_mr <- get_mr(study = "deco", dir = "/scratch/groups/ukbiobank/usr/alish/AD_MR/", extra = "ADexp/")

```

# Combine all to count how many exposures we had as proteins
```{r echo=FALSE}
# combine all to count how many exposures and outcomes we had
mr_df_all <- rbind(folk_mr, suhre_mr, sun_mr, wood_mr, ahol_mr, scal_mr, hill_mr, hogl_mr, enroth_mr, deco_mr)
als <- str_count(mr_df_all$Exposure, "AD")
als[is.na(als)] <- 0
sum(als)
```

# Combine all to count how many exposures we had as ALS
```{r}
markers <- mr_df_all[mr_df_all$Exposure %like% "AD_Schwartzentruber", ]

nrow(markers)
```

# Combine full data, work out odds ratios and 95% CI
```{r}

mr_df_all$Beta_exponent <- exp(mr_df_all$bxy)
mr_df_all$LCI <- mr_df_all$Beta_exponent - (mr_df_all$se * 1.96)
mr_df_all$UCI <- mr_df_all$Beta_exponent + (mr_df_all$se * 1.96)

mr_df_all$Exposure <- gsub(".txt", "",mr_df_all$Exposure)
mr_df_all$Outcome <- gsub(".txt", "",mr_df_all$Outcome)
```

# Compute q-values and add them to the large combined data frame
```{r}
pvalues <- mr_df_all$p
qobj <- qvalue(p = pvalues)
lfdr <- qobj$lfdr
summary(qobj)
hist(qobj)
plot(qobj)

qvalues <- as.matrix(qobj$qvalues)
mr_df_all$qvalue <- cbind(qvalues)
```


# Adjusted p-values using SGoF method
```{r}
p <- SGoF(mr_df_all$p)
summary(p)

plot(p)

p_list <- p$Adjusted.pvalues

mr_df_all <- mr_df_all[order(mr_df_all$p),]
mr_df_all$sgof <- p_list
```

# All effects AD exposure
```{r}

mr_df_all$p.adjust <- p.adjust(p = mr_df_all$p, method = "fdr")

AD_exposure <- mr_df_all[mr_df_all$Exposure %like% "AD_Schwartzentruber", ]

  AD_exposure %>%
  kbl() %>%
  kable_classic_2(full_width = F)

```

# Sort by significant p-values
```{r}
mr_df_filtered <- AD_exposure %>%
  filter(qvalue < 0.05)

mr_df_filtered <- mr_df_filtered %>%
  rename(Beta = bxy, StandardError = se, Pvalue = p, SNPs = nsnp, LowerCI = LCI, UpperCI =  UCI, AdjustedPvalue = p.adjust)

mr_df_filtered <- mr_df_filtered %>% arrange(desc(Beta_exponent))

mr_df_filtered$Log_Beta <- log(mr_df_filtered$Beta_exponent)
mr_df_filtered$Log_LowerCI <- log(mr_df_filtered$LowerCI)
mr_df_filtered$Log_UpperCI <- log(mr_df_filtered$UpperCI)

  mr_df_filtered %>%
  kbl() %>%
  kable_classic_2(full_width = F)

```

# Create a figure with AD as the exposure

```{r echo=FALSE}

mr_df_chart <- mr_df_filtered %>%
  filter(qvalue < 0.0001)

AD_exposure_stringent <- mr_df_chart
vroom_write(x = AD_exposure_stringent, file = "/scratch/groups/ukbiobank/usr/alish/AD_MR/AD_exposure_stringent")

mr_df_chart <- mr_df_chart[complete.cases(mr_df_chart), ]

#Tidy up the names of some proteins
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)
mr_df_chart$Exposure <- sub("^(.*)[.].*", "\\1",mr_df_chart$Exposure)

mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)
mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)
mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)
mr_df_chart$Outcome <- sub("^(.*)[.].*", "\\1",mr_df_chart$Outcome)

mr_df_chart <- mr_df_chart[!duplicated(mr_df_chart$Outcome), ]

library(scales)

ggplot(mr_df_chart, aes(x=reorder(as.factor(Outcome), Beta_exponent) , y=Beta_exponent)) +
    geom_boxplot(fill="slateblue") + coord_flip() + geom_errorbar(aes(ymax = UpperCI, ymin = LowerCI))+ ggtitle("Exposure - Alzheimer's") +
  xlab("Outcome Variable") + ylab("Odds Ratio (95% Confidence Interval)")+ geom_hline(yintercept=1, linetype="dashed", color = "red") + ylim(-0.5, 2.5) +
  theme_bw() + theme(text=element_text(size=10, family="Arial")) +
    theme(panel.grid.major.x = element_line(size = 0.5,
                                        linetype = 'dashed',
                                        colour = "gray41"),
        panel.grid.major.y = element_line(size = 0.1,
                                        linetype = 'solid',
                                        colour = "gray62")) +
  scale_y_log10(breaks = c(-1, 0.5, 1, 1.5, 2, 2.5)) +
  guides(color = guide_legend(reverse = TRUE)) +
  coord_flip(clip = "off", ylim = c(-1,3))

```
